---
title: "Introduction"
format: revealjs
---

## Workspace Hygiene

. . .

**Start from a clean slate!**

. . .


Your workspace is your laboratory - keep it free of contamination!

. . .

***Tools > Global Options > General***

In the "Basic" tab

- Unceck "Restore .RData into workspace at startup"
- Set "Save workspace to .RData on exit" to "Never"

## Absolute and Relative Paths

. . .

Your code should run on any machine!

. . .

```{r}
#| eval: false
#| echo: true
# BAD PRACTICE: path and data does not exist on other machines
read.csv("C:/path/to/important/raw/data.csv")
```

. . .

```{r}
#| eval: false
#| echo: true
# BAD PRACTICE: works only on one specific machine
setwd("C:/path/to/folder/that/only/exists/on/my/machine")
```

. . .

```{r}
#| eval: false
#| echo: true
# BAD PRACTICE: requires manual work
# to run this script, go to
# Session -> Set Working Directory -> To Source File Location
# to set your working directory correctly
```

## Introducing: RProjects

::: {.incremental}

- Project files that set working directory automatically after opening the project
- Whole project folder can be passed between machines and people
- Specific to RStudio

:::

## Starting with RProjects

:::: {.columns}

::: {.column width="50%"}

File > New Project

Then select your preferred option

:::

::: {.column width="50%"}
Picture coming soon

:::

::::

## Working with RProjects


:::: {.columns}

::: {.column width="50%"}

File > Open Project > Select RProject file

Work in Rstudio as usual

All paths are set relative to the .RProject file

:::

::: {.column width="50%"}
Picture coming soon

:::

::::



